# Caluclate TE:KRAB-ZNF from SQuIRE output

From previous script: `squire.qmd`, I obtain 27 mapping and quantifying files.
Here I demonstrate how to calculate their correlations using TEKRABber.
Note: this file is first duplicated from rmcorr_human.qmd. Thus, some variables 
look like human sample.

```{r message=FALSE}
library(dplyr)
library(TEKRABber)
library(twice)
library(ggplot2)
library(tidyr)
library(rmcorr)
library(doParallel)
library(GenomicFeatures)
library(biomaRt)

data("hmKZNFs337")
```

Read expression file, read TE files:

```{r}
# human 9 sample
path="~/github/kap1_primate/bcells_exp/squire_count/chimp_count/"
# Judith
h1 <- read.table(paste0(path,"Judith_count/Chimp_Judith_Zeb2_neg_R1_TEcounts.txt"), header = T)
h2 <- read.table(paste0(path,"Judith_2/Chimp_Judith_Zeb2_2_R1.fastq_TEcounts.txt"), header = T)
h3 <- read.table(paste0(path,"Judith_3/Chimp_Judith_Zeb2_3_R1.fastq_TEcounts.txt"), header = T)

# Leo
h4 <- read.table(paste0(path,"Leo_count/Chimp_Leo_Zeb2_neg_R1_TEcounts.txt"), header = T)
h5 <- read.table(paste0(path,"Leo_2/Chimp_Leo_Zeb2_2_R1_TEcounts.txt"), header = T)
h6 <- read.table(paste0(path,"Leo_3/Chimp_Leo_Zeb2_3_R1_TEcounts.txt"), header = T)

# Maryke
h7 <- read.table(paste0(path,"Maryke_count/Chimp_Maryke_Zeb2_neg_R1_TEcounts.txt"), header = T)
h8 <- read.table(paste0(path,"Maryke_2/Chimp_Maryke_Zeb2_2_R1.fastq_TEcounts.txt"), header = T)
h9 <- read.table(paste0(path,"Maryke_3/Chimp_Maryke_Zeb2_3_R1.fastq_TEcounts.txt"), header = T)

filter_select <- function(df, sample){
    
    # 1. preprocess
    df_process <- df %>%
        filter(score >= 90) %>%
        mutate(len = TE_stop - TE_start) %>%
        dplyr::select(c(4,17,19))
    colnames(df_process) <- c("ID", sample, "length")
    df_process <- df_process[!duplicated(df_process$ID), ]
    
    # 2. calculate RPK
    df_process <- df_process %>%
        mutate(RPK = !!sym(sample) / (length / 1000))  # Calculate Reads Per Kilobase
    
    # 3. calculate TPM
    rpk_sum <- sum(df_process$RPK)  # Sum of all RPKs
    df_process <- df_process %>%
        mutate(TPM = RPK / rpk_sum * 1e6)  # Normalize RPK to TPM
    
    # Return the processed dataframe with TPM
    df_process_result <- df_process[,c(1,5)]
    colnames(df_process_result) <- c("ID", sample)
    
    return(df_process_result)
}

h1_f <- filter_select(h1, "c1")
h2_f <- filter_select(h2, "c2")
h3_f <- filter_select(h3, "c3")
h4_f <- filter_select(h4, "c4")
h5_f <- filter_select(h5, "c5")
h6_f <- filter_select(h6, "c6")
h7_f <- filter_select(h7, "c7")
h8_f <- filter_select(h8, "c8")
h9_f <- filter_select(h9, "c9")

# merge
df_h_TE <- Reduce(function(x, y) merge(x, y, by = "ID", all = FALSE), 
                    list(h1_f, h2_f, h3_f, h4_f, h5_f, h6_f, h7_f, h8_f, h9_f)) #3006
rownames(df_h_TE) <- paste0("chimp", 1:nrow(df_h_TE))
```

For chimpanzee KRAB-ZNFs,
```{r}
g_path <- "~/github/kap1_primate/bcells_exp/TEtranscript/chimp_counts/"
#Judith
g1 <- read.table(paste0(g_path,"Chimp_Judith_Zeb2_neg.cntTable"), header=TRUE)
g2 <- read.table(paste0(g_path,"Chimp_Judith_Zeb2_2.cntTable"), header=TRUE)
g3 <- read.table(paste0(g_path,"Chimp_Judith_Zeb2_3.cntTable"), header=TRUE)

#Leo
g4 <- read.table(paste0(g_path,"Chimp_Leo_Zeb2_neg.cntTable"), header=TRUE)
g5 <- read.table(paste0(g_path,"Chimp_Leo_Zeb2_2.cntTable"), header=TRUE)
g6 <- read.table(paste0(g_path,"Chimp_Leo_Zeb2_3.cntTable"), header=TRUE)

#Maryke
g7 <- read.table(paste0(g_path,"Chimp_Maryke_Zeb2_neg.cntTable"), header=TRUE)
g8 <- read.table(paste0(g_path,"Chimp_Maryke_Zeb2_2.cntTable"), header=TRUE)
g9 <- read.table(paste0(g_path,"Chimp_Maryke_Zeb2_3.cntTable"), header=TRUE)
```

```{r}
g_list <- list(g1=g1, g2=g2, g3=g3, g4=g4, g5=g5, g6=g6, g7=g7, g8=g8, g9=g9)
merged_g <- Reduce(function(x,y) full_join(x, y, by="gene.TE"), g_list)
colnames(merged_g)[-1] <- names(g_list)
filtered_g <- merged_g %>% filter(grepl("^ENSPTRG", gene.TE))
colnames(filtered_g)[1] <- "ENSID"

# create gene length file
txdb <- makeTxDbFromGFF("~/Downloads/panTro4.ensGene.gtf")
gene_length <- genes(txdb) %>%
    as.data.frame() %>%
    mutate(length = width) %>%
    dplyr::select(ENSID=gene_id, length)
rownames(gene_length) <- 1:nrow(gene_length)
```

We need additional step to convert ENSPTRG to ENSG

```{r}
ref <- useEnsembl("ensembl", dataset = "ptroglodytes_gene_ensembl", version=105)
compare <- useEnsembl("ensembl", dataset = "hsapiens_gene_ensembl", version=105)

orthologTable <- getLDS(
    attributes = c("ensembl_gene_id"), mart = ref,
    attributesL = c("ensembl_gene_id"), martL = compare
)

```

```{r}
# add gene length information to expression
filtered_g <- filtered_g %>%
    left_join(gene_length, by="ENSID")

# calculate TPM
calculate_tpm <- function(counts, lengths) {
  rpk <- counts / (lengths / 1000)  # RPK calculation
  scaling_factor <- sum(rpk) / 1e6  # Scaling factor for TPM
  tpm <- rpk / scaling_factor       # TPM calculation
  return(tpm)
}

# convert raw counts to TPM
filtered_g <- filtered_g %>%
    mutate(across(starts_with("g"), ~ calculate_tpm(.x, length)))

# convert ENSPTRG to ENSG
filtered_g$ENSID <- substr(filtered_g$ENSID, 1, 18)
filtered_g <- filtered_g %>%
    left_join(orthologTable, join_by(ENSID==Gene.stable.ID))
```


```{r}
# select genes for correlation analysis
g_select <- rbind(hmKZNFs337, c("ENSG00000130726", "TRIM28"))

# need the ortholog information
filtered_g <- filtered_g %>%
    inner_join(g_select, join_by(Gene.stable.ID.1==ensembl_gene_id)) #269
    
rownames(filtered_g) <- filtered_g$external_gene_name
filtered_g_nozero <- filtered_g %>%
    filter(rowSums(.==0) < 3) %>%
    dplyr::select(c(2:10)) #238
```


## after obtaining dataframe
input of TE: `df_h_TE`, 
input of KRAB-ZNFs: `df_k_merge_nozero`
the function I am going to use is `src/corrOrthologTE_rmcorr.R`
```{r}
te589 <- read.table("../../tables/TE_589_intersect.csv", sep=',', header=TRUE)

# separate TE ID to a new column for filtering data
df_h_TE_filter <- df_h_TE %>% 
    mutate(teName = sapply(
        stringr::str_split(ID, "\\|"), 
        function(x) stringr::str_extract(x[4], "^[^:]+"))) %>%
    filter(teName %in% te589$X) #6096


c_sample_meta <- paste0("C", rep(1:3, each=3))

# prepare input format
df_h_TE_input <- df_h_TE_filter[,c(2:10)]
rownames(df_h_TE_input) <- df_h_TE_filter$ID

Sys.time()
chimp_rmcorr <- corrOrthologTE_rmcorr(
    geneInput = filtered_g_nozero,
    teInput = df_h_TE_input,
    numCore = 3,
    sample_meta = c_sample_meta
)
Sys.time() #25min
```


```{r}
chimp_rmcorr_addTE <- chimp_rmcorr %>%
    left_join(df_h_TE_filter[,c(1,11)], join_by(teName==ID))

chimp_rmcorr_addTE_sig <- chimp_rmcorr_addTE %>%
    filter(pvalue<0.01) #39136, lowest padj 0.2

write.csv(chimp_rmcorr_addTE_sig, file="../../tables/chimp_rmcorr_result.csv")
```

