# Analyze if genes and TEs are on the same strand

-   Input:

    -   Genes detected from GO terms (**Table S4-1**, **S4-2**, **S4-3**)

        -   including information of start, end, and strand position.

    -   TEs from reference

        -   including information of their start, end, and strand position.

-   Output:

    -   Overlap of Genes and TEs with their strand information \## load library

    ```{r}
    library(dplyr)
    library(data.table)
    library(GenomicRanges)
    library(ggplot2)
    library(patchwork)

    # remeber to source script src/analyze_gene_te_peak_interaction.R
    ```

## 1 Read input data

```{r}
# path of data
# human peak
hm_peak <- read.table("/Users/yaochung41/Desktop/phd/kap1_primate/peaks/human_annotated_peaks.txt", sep='\t', header=T)
chimp_peak <- read.table("/Users/yaochung41/Desktop/phd/kap1_primate/peaks/chimp_annotated_peaks.txt", sep='\t', header=T)
oran_peak <- read.table("/Users/yaochung41/Desktop/phd/kap1_primate/peaks/oran_annotated_peaks.txt", sep='\t', header=T)

# gene name
hm_gene <- read.table("~/Downloads/TableS4-1.tsv", sep='\t', header=T)
chimp_gene <- read.table("~/Downloads/TableS4-2.tsv", sep='\t', header=T)
oran_gene <- read.table("~/Downloads/TableS4-3.tsv", sep='\t', header=T, quote="", fill=TRUE, comment.char="")

# gene reference
# HGNC track
#hm_gene_ref <- fread("~/Downloads/hg19_gene_reference.tsv")
hm_gene_ref <- fread("~/Downloads/hg19_gene_reference.tsv")
hm_gene_ref <- hm_gene_ref %>% select(c("ensembl_gene_id", '#chrom', "strand", "chromStart", "chromEnd"))

# chimp data
chimp_gene_ref <- fread("~/Downloads/panTro4_gene_reference.tsv")
chimp_map      <- fread("~/Downloads/panTro4_TtoG.tsv")
chimp_ref_final <- merge(
  chimp_gene_ref[, .(transcript = name, `#chrom` = chrom, strand, chromStart = txStart, chromEnd = txEnd)], 
  chimp_map[, .(transcript, gene_id = `#gene`)], 
  by = "transcript", 
  all.x = TRUE
)[, `:=`(
  ensembl_gene_id = substr(gene_id, 1, 18), # Clean ID
  transcript = NULL,                        # Remove temporary transcript ID
  gene_id = NULL                            # Remove temporary gene_id
)]

# orangutan
oran_gene_ref <- fread("~/Downloads/ponAbe2_gene_reference.tsv")
oran_map      <- fread("~/Downloads/ponAbe2_TtoG.tsv")

oran_ref_final <- merge(
  oran_gene_ref[, .(transcript = name, `#chrom` = chrom, strand, chromStart = txStart, chromEnd = txEnd)], 
  oran_map[, .(transcript, gene_id = `#gene`)], 
  by = "transcript", 
  all.x = TRUE
)[, `:=`(
  ensembl_gene_id = substr(gene_id, 1, 18), # Convert to Gene ID and clean to 18 chars
  transcript = NULL,                        # Cleanup temp columns
  gene_id = NULL
)]


# TE reference
hm_TE_ref <- fread("~/Downloads/hg19_TE_reference.tsv")
chimp_TE_ref <- fread("~/Downloads/panTro4_TE_reference.tsv")
oran_TE_ref <- fread("~/Downloads/ponAbe2_TE_reference.tsv")
```

Test

```{r}
human_result <- analyze_gene_te_peak_intersection(
    peak_data = hm_peak,
    gene_list = hm_gene,
    gene_ref = hm_gene_ref,
    te_ref = hm_TE_ref
)
```

```{r}
chimp_result <- analyze_gene_te_peak_intersection(
    peak_data = chimp_peak,
    gene_list = chimp_gene,
    gene_ref = chimp_ref_final,
    te_ref = chimp_TE_ref
)
```

```{r}
oran_result <- analyze_gene_te_peak_intersection(
    peak_data = oran_peak,
    gene_list = oran_gene,
    gene_ref = oran_ref_final,
    te_ref = oran_TE_ref
)
```

```{r}
# save table
write.table(human_result$data, "tables/human_gene_TE_strand.tsv", sep='\t', quote=FALSE, row.names=FALSE, col.names=TRUE)
write.table(chimp_result$data, "tables/chimp_gene_TE_strand.tsv", sep='\t', quote=FALSE, row.names=FALSE, col.names=TRUE)
write.table(oran_result$data, "tables/oran_gene_TE_strand.tsv", sep='\t', quote=FALSE, row.names=FALSE, col.names=TRUE)

```
At this stage, we can know use a table to write the number of antisense and sense:

|                |           |                |
|----------------|-----------|----------------|
|                | **Sense** | **Anti-Sense** |
| **human**      | 25        | 26             |
| **chimpanzee** | 357       | 452            |
| **orangutan**  | 344       | 528            |

Next, calculate the distribution of TE subfamily that are related to antisense and sense.
Remeber to source `analyze_te_strand_distribution.R`

```{r}         
# add back repFamily and repClass information
# human
human_class_res <- analyze_te_strand_distribution(human_result$data, hm_TE_ref, "Human", group_by_var = "repClass")
print(human_class_res$plot)

human_family_res <- analyze_te_strand_distribution(human_result$data, hm_TE_ref, "Human", group_by_var = "repFamily", min_count = 1)
print(human_family_res$plot)

human_family_pref <- analyze_te_strand_preference(human_family_res$extended_data, group_by_var = "repFamily")
head(human_family_pref)

```

```{r}
# add back repFamily and repClass information
# chimpanzee
chimp_class_res <- analyze_te_strand_distribution(chimp_result$data, chimp_TE_ref, "Chimpanzee", group_by_var = "repClass")
print(chimp_class_res$plot)

chimp_family_res <- analyze_te_strand_distribution(chimp_result$data, chimp_TE_ref, "Chimpanzee", group_by_var = "repFamily", min_count = 1)
print(chimp_family_res$plot)

chimp_family_pref <- analyze_te_strand_preference(chimp_family_res$extended_data, group_by_var = "repFamily")
head(chimp_family_pref)
```


```{r}
# add back repFamily and repClass information
# orangutan
oran_class_res <- analyze_te_strand_distribution(oran_result$data, oran_TE_ref, "Orangutan", group_by_var = "repClass")
print(oran_class_res$plot)

oran_family_res <- analyze_te_strand_distribution(oran_result$data, oran_TE_ref, "Orangutan", group_by_var = "repFamily", min_count = 1)
print(oran_family_res$plot)

oran_family_pref <- analyze_te_strand_preference(oran_family_res$extended_data, group_by_var = "repFamily")
head(oran_family_pref)
```

Save detection

```{r}
write.table(human_family_pref, "tables/human_gene_TE_strand_detection.tsv", sep='\t', quote=FALSE, row.names=FALSE, col.names=TRUE)
write.table(chimp_family_pref, "tables/chimpanzee_gene_TE_strand_detection.tsv", sep='\t', quote=FALSE, row.names=FALSE, col.names=TRUE)
write.table(oran_family_pref, "tables/oran_gene_TE_strand_detection.tsv", sep='\t', quote=FALSE, row.names=FALSE, col.names=TRUE)

```

```{r}
combined_plot <- (
  (human_class_res$plot | human_family_res$plot) /
  (chimp_class_res$plot | chimp_family_res$plot) /
  (oran_class_res$plot  | oran_family_res$plot)
) + 
  plot_layout(guides = 'collect') & # merge repeat plot
  theme(legend.position = "bottom")

print(combined_plot)
```


```{r}
combined_plot <- combined_plot + 
  plot_annotation(
    title = "Comparative Analysis of TE Strand Detection",
    subtitle = "Class and Family Level",
    tag_levels = 'A' # add A, B, C, D, E, F tag for figures
  )

print(combined_plot)
ggsave("figures/TE_Strand_Comparison.pdf", combined_plot, width = 12, height = 18)
```