# Prepare input data for MEME-ChIP analysis (v5.5.7)

```{r}
library(dplyr)
```


```{r}
hm <- read.table("../../peaks/macs3_results/human_TE_only.bed", sep="\t") #2077
chimp <- read.table(
    "../../peaks/macs3_results/chimp_TE_only.bed", sep="\t") #5754
oran <- read.table("../../peaks/macs3_results/oran_TE_only.bed", sep="\t") #10098
```

I upload the human_TE_only.bed, chimp_TE_only.bed, and oran_TE_only.bed files to MEME-ChIP platform. 

https://meme-suite.org/meme/info/status?service=MEMECHIP&id=appMEMECHIP_5.5.717321133616821376774291
(link will be expired after 24th of Nov.)

Then I can download the data from MEME result: `kap1_primate/meme_result`.
The purpose is to get the motifs information from KRAB-ZNFs.

```{r}
human_summary <- read.table('../../meme_result/MEME_human/summary.tsv', 
                            sep='\t', header=TRUE)

chimp_summary <- read.table('../../meme_result/MEME_chimp/summary.tsv',
                            sep='\t', header=TRUE)

oran_summary <- read.table('../../meme_result/MEME_oran/summary.tsv',
                           sep='\t', header=TRUE)
```

Get rows with KRAB-ZNF motif.
```{r}
hm_select <- human_summary[
    apply(human_summary, 1, function(x) any(grepl("ZNF|TRIM28|PRDM9|ZKSCAN", x))), ] #36

chimp_select <- chimp_summary[
    apply(chimp_summary, 1, function(x) any(grepl("ZNF|TRIM28|PRDM9|ZKSCAN", x))), ] #2

oran_select <- oran_summary[
    apply(oran_summary, 1, function(x) any(grepl("ZNF|TRIM28|PRDM9|ZKSCAN", x))), ] #11

```

merge dataframe and upload to supplementary tables
```{r}
df <- hm_select %>%
    select(c(2:12)) %>%
    mutate(species = "human")

chimp_process <- chimp_select %>%
    select(c(2:12)) %>%
    mutate(species = "chimp")
    
oran_process <- oran_select %>%
    select(c(2:12)) %>%
    mutate(species = "orangutan")

# rbind
df_motif <- rbind(df, chimp_process, oran_process)

df_motif <- df_motif[,c(12, seq(1:11))]
rownames(df_motif) <- 1:nrow(df_motif)
write.table(df_motif, file="../../tables/motifs.csv", sep=',', row.names = FALSE)
```


